Non-convergence in iterative imputation

### 2025-04-10 Manuscript/manuscript.qmd
add complete data to sim

```{r setup, echo=FALSE}
library(mice)
library(ggmice)
library(ggplot2)
```

@@

All inferences rely on the convergence of the imputation algorithm, yet determining whether an algorithm has converged is not trivial. There has not been a systematic study on how to evaluate the convergence of iterative imputation algorithms. A widely accepted practice is visual inspection of the algorithm, although diagnosing convergence through visual inspection may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18]. Therefore, a quantitative diagnostic method to assess convergence would be preferred.

@@

We use empirical incomplete data: the `boys` dataset from the `mice` package, which contains data on some health-related variables for 748 Dutch boys [@mice]. Say we're interested in the relation between children's heights and their respective ages, we could use a linear regression model to predict `hgt` from `age`. However, as figure XYZ shows, the variable `hgt` is not completely observed. To be able to analyze these data, we need to solve the missing data problem first.

```{r boys, echo=FALSE}
ggmice::ggmice(mice::boys, ggplot2::aes(age, hgt)) + ggplot2::geom_point()
```

@@

```{r trace, echo=FALSE}
imp <- mice::mice(mice::boys[, c("age", "hgt", "wgt")], maxit = 20, printFlag = FALSE)
plot_trace(imp)
```

-   Conventional inspection of algorithmic

-   We are interested in predicting age from the other variables, in particular in the regression coefficient of `hgt`

-   We compare non-convergence identified using visual inspection versus rhat in the chain variances, scientific estimate and lambda.

-   The figures show results of a `mice` run with 20 iterations but otherwise default settings.

```{r case, fig.cap = "Case study.", echo=FALSE, eval=FALSE}
trace <- param %>% 
  ggplot() + 
  geom_line(aes(x = it, y = mu.hgt, color = as.factor(m))) + 
  theme_classic() + 
  labs(x = "Iteration number",
       y = "Chain mean variable 'hgt'",
       color = "Imputation")
means <- casestudy_plot(v="r.hat.max.mu.hgt", name = "Rhat chain means") + 
  list(geom_hline(yintercept = 1.2, color = "grey"))
qhats <-
  casestudy_plot(v = "r.hat.max.qhat", name = "Rhat regression estimate") +
  list(geom_hline(yintercept = 1.2, color = "grey"))
lambdas <- casestudy_plot(v = "r.hat.max.lambda", name = "Rhat lambda") +
  list(geom_hline(yintercept = 1.2, color = "grey"))
(trace + means) / (qhats + lambdas) + plot_layout(guides="collect") + plot_annotation(tag_levels = 'A')

```

@@

This figure (\@ref(fig:case)B) shows that 7 iterations are required before the $\widehat{R}$-values of the chain means drop below the threshold for non-convergence.

The $\widehat{R}$-values for the scientific estimate reaches the threshold much sooner, when $n_{it}=14$ (see \@ref(fig:case)C).

According to the $\widehat{R}$-values with $\lambda$ as parameter, at least 15 iterations are required (see \@ref(fig:case)D).

```{r}

```

@@

The potential scale reduction factor metric assumes over-dispersed starting values of the iterative algorithm. The MICE algorithm does not start in an over-dispersed state.  MICE does not rely on staring values for parameters at all: instead, the algorithm is initialized based on a 'zeroth' iteration, where missing values are filled in using a random draw from observed values. Under MCAR, this would typically not yield an over-dispersed state at all, depending on the parameter of interest: means and variances are not affected, regression coefficients and other multivariate parameters start biased downwards (because sampling starting values distorts multivariate structures in the data). Whether this accrues to over-dispersion is questionable. Under MAR (or MNAR), some over-dispersion might occur when the observed data and imputed data differ impactfully. The algorithm should 'escape' the initial state (based on the staring values drawn from observed data alone). Over iterations, all parameters of interest (e.g. means, variances, multivariate estimands) should converge towards a stable state. Moreover, the MICE algorithm is initialized with more information than a typical MCMC algorithm: parameter estimates depend not only on imputed data, but also on observed data that does not change over iterations. Therefore, the initial state of the algorithm is 'too good' to observe a relative decrease in $\widehat{R}$.

>>>>

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(mice)
library(ggmice)
library(ggplot2)
```

@@

All inferences with imputed data rely on the convergence of the imputation algorithm. Yet, determining whether an algorithm has converged is not trivial. There has not been a systematic study on how to evaluate the convergence of iterative imputation algorithms. A widely accepted practice is visual inspection of the algorithm. Diagnosing convergence through visual inspection, however, may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18]. Therefore, a quantitative diagnostic method to assess convergence would be preferred.

@@

We use empirical incomplete data: the `boys` dataset from the `mice` package, which contains health-related data for 748 Dutch boys [@mice]. Say, we're interested in the relation between children's heights and their respective ages, we could use a linear regression model to predict `hgt` from `age`. However, as figure XYZ shows, the variable `hgt` is not completely observed. To be able to analyze these data, we need to solve the missing data problem first.

```{r boys, echo=FALSE}
ggmice::ggmice(mice::boys, ggplot2::aes(age, hgt, shape = .where)) + 
  ggplot2::geom_point() + 
  ggplot2::scale_shape_manual(values = c(1, 4)) +
  ggplot2::labs(x = "Age (years)", y = "Height (cm)", shape = "")
```

@@

```{r trace, echo=FALSE}
imp <- mice::mice(mice::boys[, c("age", "hgt", "wgt")], maxit = 20, printFlag = FALSE)
plot_trace(imp, "hgt") 
```

```{r}
# impute missingness with mis-specified imputation model
meth <- make.method(boys) #define imputation model
meth["bmi"] <- "~I(wgt / (hgt / 100)^2)" #mis-specify model
nonconv <- mice( #impute missingness
  boys,
  meth = meth,
  maxit = 20,
  print = FALSE,
  seed = 60109
)

# # save results
# save(nonconv, file = "Results/example_nonconv.Rdata")

#####################
# Regular convergence
#####################

# impute missingness with correctly specified imputation model
pred <- make.predictorMatrix(boys) #mend imputation method
pred[c("hgt", "wgt"), "bmi"] <- 0 #remove dependency
conv <- mice( #impute missingness
  boys,
  meth = meth,
  pred = pred,
  maxit = 20,
  print = FALSE,
  seed = 60109
)

plot_trace(nonconv, "hgt") +
  ggplot2::labs(title = "Pathological non-convergence")
plot_trace(conv, "hgt") +
  ggplot2::labs(title = "Typical convergence")
```

@@

## Aims

- can we quantify non-convergence quantitatively?
- which parameter should we track?
- how many iterations are needed before convergence is reached?
- how many iterations are needed before valid inferences are reached?

@@

The potential scale reduction factor metric assumes over-dispersed starting values of the iterative algorithm. The MICE algorithm does not start in an over-dispersed state.  MICE does not rely on starting values for parameters at all: instead, the algorithm is initialized based on a 'zeroth' iteration, where missing values are filled in using a random draw from observed values. Under MCAR, this would typically not yield an over-dispersed state at all, depending on the parameter of interest: means and variances are not affected, regression coefficients and other multivariate parameters start biased downwards (because sampling starting values distorts multivariate structures in the data). Whether this accrues to over-dispersion is questionable. Under MAR (or MNAR), some over-dispersion might occur when the observed data and imputed data differ impactfully. The algorithm should 'escape' the initial state (based on the starting values drawn from observed data alone). Over iterations, all parameters of interest (e.g. means, variances, multivariate estimands) should converge towards a stable state. Moreover, the MICE algorithm is initialized with more information than a typical MCMC algorithm: parameter estimates depend not only on imputed data, but also on observed data that does not change over iterations. Therefore, the initial state of the algorithm is 'too good' to observe a relative decrease in $\widehat{R}$.

### 2025-04-10 Manuscript/template.Rmd
add complete data to sim

```{r data}
performance <- readRDS("./Results/performance.RDS")
summary <- performance %>% 
  group_by(method, mech, prop, .it, term) %>% 
  summarise(across(c(bias, cov, ciw, ac, psrf), mean, na.rm = TRUE))
SEs <- performance %>% 
  group_by(method, mech, prop, .it, term) %>% 
  summarise(across(c(bias, ciw, ac, psrf), var, na.rm = TRUE)) %>% 
  mutate(across(c(bias, ciw, ac, psrf), sqrt))
results <- left_join(summary, SEs, by = c("method", "mech", "prop", ".it", "term"), suffix = c("", "_SE")) %>% 
  mutate(cov = cov,
         cov_SE = sqrt((cov * (1 - cov)) / 2000))
```

@@

results %>% 
  filter(term == coeff, .it > 0, .it < maxit + 1) %>% 
  ggplot(aes(x = .it, y = ac)) +
  geom_hline(yintercept = 0, color = "black") +
  geom_errorbar(aes(ymin = ac - ac_SE, ymax = ac + ac_SE), alpha = 0.1) +
  geom_line(color = "blue") +
  geom_point(color = "blue") +
  facet_grid(as.factor(prop) ~ mech) +
  theme_classic()

results %>% 
  filter(term == coeff, .it > 0, .it < maxit + 1) %>% 
  ggplot(aes(x = .it, y = psrf)) +
  geom_hline(yintercept = 1, color = "black") +
  geom_errorbar(aes(ymin = psrf - psrf_SE, ymax = psrf + psrf_SE), alpha = 0.1) +
  geom_line(color = "blue") +
  geom_point(color = "blue") +
  facet_grid(as.factor(prop) ~ mech) +
  theme_classic()
```

>>>>

```{r data}
performance <- readRDS("./Results/performance.RDS")
summary <- performance %>% 
  group_by(method, mech, prop, .it, term) %>% 
  summarise(across(c(bias, cov, ciw, ac_mean, psrf_mean, ac_sd, psrf_sd), mean, na.rm = TRUE))
SEs <- performance %>% 
  group_by(method, mech, prop, .it, term) %>% 
  summarise(across(c(bias, cov, ciw, ac_mean, psrf_mean, ac_sd, psrf_sd), var, na.rm = TRUE)) %>% 
  mutate(across(c(bias, cov, ciw, ac_mean, psrf_mean, ac_sd, psrf_sd), sqrt))
results <- left_join(summary, SEs, by = c("method", "mech", "prop", ".it", "term"), suffix = c("", "_SE")) %>% 
  mutate(cov = cov,
         cov_SE = sqrt((cov * (1 - cov)) / 2000))
```

@@

results %>% 
  filter(term == coeff, .it > 0, .it < maxit + 1) %>% 
  ggplot(aes(x = .it, y = ac_mean)) +
  geom_hline(yintercept = 0, color = "black") +
  geom_errorbar(aes(ymin = ac_mean - ac_mean_SE, ymax = ac_mean + ac_mean_SE), alpha = 0.1) +
  geom_line(color = "blue") +
  geom_point(color = "blue") +
  facet_grid(as.factor(prop) ~ mech) +
  theme_classic()

results %>% 
  filter(term == coeff, .it > 0, .it < maxit + 1) %>% 
  ggplot(aes(x = .it, y = psrf_mean)) +
  geom_hline(yintercept = 1, color = "black") +
  geom_errorbar(aes(ymin = psrf_mean - psrf_mean_SE, ymax = psrf_mean + psrf_mean_SE), alpha = 0.1) +
  geom_line(color = "blue") +
  geom_point(color = "blue") +
  facet_grid(as.factor(prop) ~ mech) +
  theme_classic()

results %>% 
  filter(term == coeff, .it > 0, .it < maxit + 1) %>% 
  ggplot(aes(x = .it, y = ac_sd)) +
  geom_hline(yintercept = 0, color = "black") +
  geom_errorbar(aes(ymin = ac_sd - ac_sd_SE, ymax = ac_sd + ac_sd_SE), alpha = 0.1) +
  geom_line(color = "blue") +
  geom_point(color = "blue") +
  facet_grid(as.factor(prop) ~ mech) +
  theme_classic()

results %>% 
  filter(term == coeff, .it > 0, .it < maxit + 1) %>% 
  ggplot(aes(x = .it, y = psrf_sd)) +
  geom_hline(yintercept = 1, color = "black") +
  geom_errorbar(aes(ymin = psrf_sd - psrf_sd_SE, ymax = psrf_sd + psrf_sd_SE), alpha = 0.1) +
  geom_line(color = "blue") +
  geom_point(color = "blue") +
  facet_grid(as.factor(prop) ~ mech) +
  theme_classic()
```
